### 2022-09-29 src/tex/ms.tex
showyourwork setup.  Let's see if it works!

% Define document class
\documentclass[twocolumn]{aastex631}
\usepackage{showyourwork}

\newcommand{\cca}{Center for Computational Astrophysics, Flatiron Institute, New York NY 10010, USA}
\newcommand{\sbu}{Department of Physics and Astronomy, Stony Brook University, Stony Brook NY 11794, USA}

% Begin!
\begin{document}

% Title
\title{Eliminating Systematic Bias in Bright-Siren Multimessenger Observations of Compact Object Mergers from Inclination-Dependent Emission}

% Author list
\author{Will M. Farr}
\email{will.farr@stonybrook.edu}
\affiliation{\sbu}
\affiliation{\cca}

% Abstract with filler text
\begin{abstract}
    This article and associated code explores systematics associated with
inclination dependence in detection of bright counterparts to GW mergers.  The
TL;DR is that, by fitting out the inclination dependence of EM detections with
fairly simple models, we can remove what otherwise would be a bias that could
prevent percent-level expansion history measurements with bright sirens.
\end{abstract}

% Main body with filler text
\section{Introduction}
\label{sec:intro}

Lorem ipsum dolor sit amet, consectetuer adipiscing elit.
Ut purus elit, vestibulum ut, placerat ac, adipiscing vitae, felis.
Curabitur dictum gravida mauris, consectetuer id, vulputate a, magna.
Donec vehicula augue eu neque, morbi tristique senectus et netus et.
Mauris ut leo, cras viverra metus rhoncus sem, nulla et lectus vestibulum.
Phasellus eu tellus sit amet tortor gravida placerat.
Integer sapien est, iaculis in, pretium quis, viverra ac, nunc.
Praesent eget sem vel leo ultrices bibendum.
Aenean faucibus, morbi dolor nulla, malesuada eu, pulvinar at, mollis ac.
Curabitur auctor semper nulla donec varius orci eget risus.
Duis nibh mi, congue eu, accumsan eleifend, sagittis quis, diam.
Duis eget orci sit amet orci dignissim rutrum.

Nam dui ligula, fringilla a, euismod sodales, sollici- tudin vel, wisi.
Morbi auctor lorem non justo, nam lacus libero, pretium at, lobortis vitae.
Donec aliquet, tortor sed accumsan bibendum, erat ligula aliquet magna.
Morbi ac orci et nisl hendrerit mollis, suspendisse ut massa, cras nec ante.
Pellentesque a nulla cum sociis natoque penatibus et magnis dis parturient.
Aliquam tincidunt urna, nulla ullamcorper vestibulum turpis.
Pellentesque cursus luctus mauris \citep{Luger2021}.

\bibliography{bib}

\end{document}

### 2022-09-29 src/tex/ms.tex
Working on text for Section 2, observational model, showing GW170817 correspondence.

@@
-% Define document class
-\documentclass[twocolumn]{aastex631}
+\documentclass[modern]{aastex631}
 \usepackage{showyourwork}
+
+\newcommand{\obs}{\mathrm{obs}}
 
@@
 
-Lorem ipsum dolor sit amet, consectetuer adipiscing elit.
-Ut purus elit, vestibulum ut, placerat ac, adipiscing vitae, felis.
-Curabitur dictum gravida mauris, consectetuer id, vulputate a, magna.
-Donec vehicula augue eu neque, morbi tristique senectus et netus et.
-Mauris ut leo, cras viverra metus rhoncus sem, nulla et lectus vestibulum.
-Phasellus eu tellus sit amet tortor gravida placerat.
-Integer sapien est, iaculis in, pretium quis, viverra ac, nunc.
-Praesent eget sem vel leo ultrices bibendum.
-Aenean faucibus, morbi dolor nulla, malesuada eu, pulvinar at, mollis ac.
-Curabitur auctor semper nulla donec varius orci eget risus.
-Duis nibh mi, congue eu, accumsan eleifend, sagittis quis, diam.
-Duis eget orci sit amet orci dignissim rutrum.
+\section{Observational Model}
 
-Nam dui ligula, fringilla a, euismod sodales, sollici- tudin vel, wisi.
-Morbi auctor lorem non justo, nam lacus libero, pretium at, lobortis vitae.
-Donec aliquet, tortor sed accumsan bibendum, erat ligula aliquet magna.
-Morbi ac orci et nisl hendrerit mollis, suspendisse ut massa, cras nec ante.
-Pellentesque a nulla cum sociis natoque penatibus et magnis dis parturient.
-Aliquam tincidunt urna, nulla ullamcorper vestibulum turpis.
-Pellentesque cursus luctus mauris \citep{Luger2021}.
+We use a simplified observational model that reproduces the essential aspects of
+gravitational wave distance measurements without requiring generating and
+inferring full gravitational waveforms.  We imagine that each gravitational wave
+event has a (cosine) inclination $x = \cos \iota$ and a distance, and these
+combine to give an amplitude in the left- and right-handed polarization
+components of 
+\begin{equation}
+    A_{R/L} = \frac{\left( 1 \pm x \right)^2}{d}
+\end{equation}
+in some units.  We assume that a detector network measures a linear projection
+of these these polarization amplitudes with unit-amplitude Gaussian noise:
+\begin{equation}
+    A_{R/L,\obs} = f_{R/L} A_{R/L} + N(0,1).
+\end{equation}
+For a detector network with two or more arbitrarily oriented, nearly
+equally-sensitive detectors, we can set $f_R = f_L = 1$; for two nearly-aligned
+equally-sensitive detectors (e.g.\ the two LIGO detectors in Washington and
+Louisiana), we can set $f_R \ll f_L = 1$ to reflect that one polarization
+component is measured much more precisely than the other.
+From here onward we assume events used for bright sirens have small localization volumes and are therefore observed in three or more detectors and we set $f_L = f_R = 1$.
+
+The S/N of an observation in our model is given by 
+\begin{equation}
+    \rho^2 = \frac{A_{R,\obs}^2 + A_{L,\obs}^2}{2}.
+\end{equation}
+We impose a Euclidean prior on $d$, $p(d) \propto d^2$, appropriate for the
+nearby universe, and a flat prior on $x$, $p(x) = 1/2$, appropriate for an
+intrinsically isotropic population of GW mergers.  Inferring $d$ and $x$ from
+$A_{R,\obs}$ and $A_{L,\obs}$ in our model results in posterior densities that
+reproduce the distance-inclination correlation observed in analysis of actual GW
+events, as well as the (fractional) distance uncertainty and inclination
+uncertainty observed in actual GW events at comparable S/N.  Figure
+\ref{fig:distance-iota} shows the inferred distance and inclination from our
+model for a GW170817-like event ($\rho \simeq 30$, $x \simeq -1$); our model's
+inferences compare directly to those for GW170817 itself \citep{Abbott2017}.
+
+\begin{figure}
+    \script{distance-iota.py}
+    \includegraphics[width=\columnwidth]{figures/distance-iota.pdf}
+    \caption{\label{fig:distance-iota} Posterior on distance $d$ and $x = \cos
+    \iota$ for a GW170817-like event ($\rho \simeq 30$) from our observational
+    model.  Contour levels correspond to succesive 10\% credible intervals; true
+    parameters are indicated by the black lines.  Compare to Figure 2 of
+    \citet{Abbott2017}.  The fractional uncertainty on $d$ after marginalizing
+    over $x$ with an isotropic (flat) prior is $\sigma_d / \left\langle d
+    \right\rangle \simeq 0.14$; for GW170817 the same quantity is also $\sigma_d
+    / \left\langle d \right\rangle \simeq \sigma_{H_0} / \left\langle H_0
+    \right\rangle \simeq 0.14$ \citep{Abbott2017}.  Similarly, our model has
+    $x_{84\%} \simeq -0.82$, while for GW170817 the equivalent 1-$\sigma$ upper
+    limit on $x \simeq -0.81$ \citep{Abbott2017}.  Our model thus does a
+    reasonable job reproducing the correlation between distance and inclination
+    and the relevant uncertainties at the S/N of GW170817; in both our model and
+    a full analysis of a GW event uncertainties will scale inversely with S/N.}
+\end{figure}
+
+\begin{acknowledgments}
+    We thank [people] for conversations.
+\end{acknowledgments}
+
+\software{\texttt{arviz} \citep{arviz_2019}, \texttt{matplotlib} \citep{Hunter:2007}, \texttt{pymc} \citep{salvatier2016probabilistic}, \texttt{seaborn} \citep{Waskom2021}, \texttt{showyourwork} \citep{Luger2021}}
 

### 2022-09-29 src/scripts/distance-iota.py

@@ -1,0 +1,21 @@
+import arviz as az
+import matplotlib.pyplot as plt
+import os.path as op
+import paths
+import seaborn as sns
+
+sns.set_theme(context='paper', style='ticks')
+
+trace = az.from_netcdf(op.join(paths.data, 'GW170817-like-chain.nc'))
+post = az.extract_dataset(trace.posterior)
+
+sns.kdeplot(data=post, x='d', y='x')
+plt.xlabel(r'$d$')
+plt.ylabel(r'$x = \cos \iota$')
+plt.tight_layout()
+plt.ylim(bottom=-1, top=-0.6)
+
+plt.axhline(-0.99, color='k')
+plt.axvline(0.13200333333333333, color='k')
+
+plt.savefig(op.join(paths.figures, 'distance-iota.pdf'))

### 2022-10-01 src/tex/ms.tex
Adding GW170817....nc to data, don't ignore data folder.
Plots committed, captions written, some text.

@@
 
+\newcommand{\todo}[1]{\textcolor{blue}{TODO: #1}}
+
+\newcommand{\dd}{\mathrm{d}}
+
+\newcommand{\detected}{\mathrm{det}}
+\newcommand{\EM}{\mathrm{EM}}
+\newcommand{\GW}{\mathrm{GW}}
 \newcommand{\obs}{\mathrm{obs}}
@@
     \includegraphics[width=\columnwidth]{figures/distance-iota.pdf}
-    \caption{\label{fig:distance-iota} Posterior on distance $d$ and $x = \cos
-    \iota$ for a GW170817-like event ($\rho \simeq 30$) from our observational
-    model.  Contour levels correspond to succesive 10\% credible intervals; true
-    parameters are indicated by the black lines.  Compare to Figure 2 of
-    \citet{Abbott2017}.  The fractional uncertainty on $d$ after marginalizing
-    over $x$ with an isotropic (flat) prior is $\sigma_d / \left\langle d
-    \right\rangle \simeq 0.14$; for GW170817 the same quantity is also $\sigma_d
-    / \left\langle d \right\rangle \simeq \sigma_{H_0} / \left\langle H_0
-    \right\rangle \simeq 0.14$ \citep{Abbott2017}.  Similarly, our model has
-    $x_{84\%} \simeq -0.82$, while for GW170817 the equivalent 1-$\sigma$ upper
-    limit on $x \simeq -0.81$ \citep{Abbott2017}.  Our model thus does a
-    reasonable job reproducing the correlation between distance and inclination
-    and the relevant uncertainties at the S/N of GW170817; in both our model and
-    a full analysis of a GW event uncertainties will scale inversely with S/N.}
+    \caption{Posterior on distance $d$ and $x = \cos \iota$ for a GW170817-like
+    event ($\rho \simeq 30$) from our observational model.  Contour levels
+    correspond to succesive 10\% credible intervals; true parameters are
+    indicated by the black lines.  Compare to Figure 2 of \citet{Abbott2017}.
+    The fractional uncertainty on $d$ after marginalizing over $x$ with an
+    isotropic (flat) prior is $\sigma_d / \left\langle d \right\rangle \simeq
+    0.14$; for GW170817 the same quantity is also $\sigma_d / \left\langle d
+    \right\rangle \simeq \sigma_{H_0} / \left\langle H_0 \right\rangle \simeq
+    0.14$ \citep{Abbott2017}.  Similarly, our model has $x_{84\%} \simeq -0.82$,
+    while for GW170817 the equivalent 1-$\sigma$ upper limit on $x \simeq -0.81$
+    \citep{Abbott2017}.  Our model thus does a reasonable job reproducing the
+    correlation between distance and inclination and the relevant uncertainties
+    at the S/N of GW170817; in both our model and a full analysis of a GW event
+    uncertainties will scale inversely with S/N.}
+    \label{fig:distance-iota}
+\end{figure}
+
+\subsection{Effect of EM Selection}
+
+We assume that each of our observations also has a detection of an
+electromagnetic counterpart that gives a perfect measurement of redshift, thus
+constraining $H_0$.  The full likelihood for both GW and EM data for an observation factorizes because the measurement noise for the observations are independent:
+\begin{equation}
+    p\left( d_\GW, d_\EM \mid x, d, H_0 \right) = p\left( d_\GW \mid x, d \right) p\left( d_\EM \mid x, d, H_0 \right).
+\end{equation}
+The first term is the likelhood representing the measurement model described
+above; the second term relates the EM measurements to the inclination, distance,
+and (through the distance-redshift relation) $H_0$.  If we were willing to model
+the dependence of the EM data on $x$, we could use it to further constrain the
+inclination measurement and thereby improve the $H_0$ measurement.  But here we
+suppose that this is not possible---either we feel that the inclination modeling
+for the EM data is systematics-dominated, or we are unable to even produce a
+model of the inclination-dependent emission being measured.  In this case we can
+marginalize the EM data out of the likelihood, retaining only (1) the redshift
+measurement and (2) the knowledge that the EM emission was detectable.  Let $\Omega$ be the set of EM data consistent with these criteria.  Then
+\begin{equation}
+    \int_{d_\EM \in \Omega} \dd d_\EM \, p\left( d_\EM \mid x, d, H_0 \right) \propto \delta\left( z - H_0 d \right) P_\mathrm{det,\EM}\left( x, d, H_0 \right),
+\end{equation}
+where $P_{\detected,\EM}\left(x, d, H_0\right)$ is the fraction of mergers with
+inclination $x$, at distance $d$, and with Hubble constant $H_0$ that produce
+detectable EM emission \citep{Mandel2019}.
+
+If we had a good model for the EM emission, we could \emph{calculate}
+$P_{\detected,\EM}$; but if we do not, we can treat it as a parameterized
+function, and learn its shape from the GW data.  (Intuitively: the GW data carry
+some information about the inclination of each event; considering the set of
+events with coincident EM detections lets us learn the population of
+inclinations for \emph{just these} events, which is proportional to
+$P_{\detected,\EM}$.)  The term $P_{\detected,\EM}$ acts like a modification of
+the prior applied to $x$, $d$, and $H_0$ in the overall analysis.  After
+integrating out the EM data in $\Omega$, we obtain a modified GW likelihood:
+\begin{multline}
+    p\left( d_\GW, z \mid x, d, H_0 \right) = \int_{d_\EM \in \Omega} \dd d_\EM p\left( d_\GW \mid x, d \right) p\left( d_\EM \mid x, d, H_0 \right) \\ \propto \delta\left( z - H_0 d \right) p\left( d_\GW \mid x, d \right) P_{\detected,\EM}\left( x, d, H_0 \right).
+\end{multline}
+
+Applying a $d^2$ prior and integrating across the $\delta$ function, we obtain 
+\begin{multline}
+    p\left( d_\GW, z \mid x, H_0 \right) \propto \int \dd d \, p\left( d_\GW, z \mid x, d, H_0 \right) d^2 \\ = \frac{z^2}{H_0^3} P_{\detected,\EM}\left( x, d = z/H_0, H_0 \right) p\left( d_\GW \mid x, d=z/H_0 \right).
+\end{multline}
+Here we see that $P_{\detected,\EM}$ acts as an additional multiplicative
+factor, appearing as a modification to the isotropic prior that we would
+naturally apply for the inclination of our gravitational wave population.
+
+\todo{Flesh this out, and discuss GW selection (mechanical, and our model, with
+$\rho > 10$); and then get specific about our model for $P_{\detected,\EM}$.
+Also discuss ``joint'' catalogs, like combining GRB-selected and
+kilonova-selected data sets.}
+
+\subsection{Functional Form of $P_{\detected,\EM}$}
+
+Here for simplicity we assume 
+\begin{itemize}
+    \item $P_{\detected,\EM}$ does not depend on $d$ wherever GW events are
+    detectable.  That is: the distance-dependence of selection effects is
+    dominated by the GW selection-function.
+    \item $P_{\detected,\EM}$ is independent of $H_0$.  This will be the case
+    whenever the choice of EM survey strategy / catalog selection does not
+    depend on the redshift of the EM sources.  For example, a flux-limited
+    survey of GW localization volumes would have a detection efficiency that is
+    independent of $H_0$; a survey that focuses on objects with
+    spectroscopically-measured redshifts consistent with the GW $d$ would have a
+    selection function that is strongly-dependent on $H_0$.
+\end{itemize}
+With these assumptions, $P_{\detected,\EM}$ is varies only with $x$.  We propose
+to use an expansion of $P_{\detected,\EM}$ in Legendre polynomials to model and
+learn the EM selection function.  Since $P_{\detected,\EM}$ must always be
+positive, we write
+\begin{equation}
+    \label{eq:Pdet-model}
+    \log P_{\detected,\EM}\left( x, d, H_0 \right) = \sum_{l=1}^{N_l} A_l P_l(x)
+\end{equation}
+(The normalization of $P_{\detected,\EM}$ does not affect the likelihood; here
+we have fixed the normalization by omitting the constant $P_0(x) = 1$ term from
+the sum.)  The $A = \left\{ A_l \mid l = 1, \ldots, N_l \right\}$ are additional
+parameters of our population model that we will infer.
+
+Some anticipated inclination-dependent selection effects fit naturally into this
+form (e.g.\ kilonovae are expected to have significant emission into all viewing
+angles \todo{citations}); others do not (e.g.\ GRB detections require a small
+viewing angle \todo{citations}).  We will show below that even for the latter
+case, where $P_{\detected,\EM}$ varies strongly---even discontinuously---with
+$x$, the smooth Legendre polynomial interpolation is sufficient to reduce
+systematics below the percent-level in a realistic measurement.
+
+\section{Examples}
+
+\subsection{Fitting Weakly Variable EM Selection}
+\label{sec:weakly-variable}
+
+Let 
+\begin{equation}
+    \label{eq:weakly-variable-efficiency}
+    \log P_{\detected,\EM}(x) = P_2(x) = \frac{1}{2}\left( 3 x^2 - 1 \right),
+\end{equation}
+and $N_l = 2$.  We generate a mock catalog of $N = 128$ GW+EM detections with
+S/N $\rho > 10$, and $H_0 = 0.7$, and calculate the joint posterior on $H_0$, $A
+= \left\{ A_1, A_2 \right\}$, and each event's $x$.  Figure
+\ref{fig:inclination-P2} shows that we recover the intrinsic inclination
+distribution of these events well.
+
+\begin{figure}
+    \script{inclination-P2.py}
+    \includegraphics[width=\columnwidth]{figures/inclination-P2.pdf}
+    \caption{The posterior on the EM detection efficiency inferred from 128 mock
+    GW+EM detections with the EM detection efficiency described in \S\
+    \ref{sec:weakly-variable}.  The blue line gives the posterior median; bands
+    give the 68\% and 95\% credible intervals.  The black line is the true
+    detection efficiency used to generate the mock data set (Eq.
+    \eqref{eq:weakly-variable-efficiency}).}
+    \label{fig:inclination-P2} 
+\end{figure}
+
+Ultimately, fitting the EM selection effects with this dataset results in
+recovery of $H_0 = \variable{output/H0-P2.txt}$ (median and 68\% credible
+interval, i.e.\ sub-percent uncertainty). Failing to account for the
+inclination-dependent EM selection effect introduces significant bias in the
+$H_0$ estimate, recovering $H_0 = \variable{output/H0-P2-nofit.txt}$
+\citep{Chen2020}.  Figure \ref{fig:H0-P2} illustrates these effects.
+
+\begin{figure}
+    \script{H0-P2.py}
+    \includegraphics[width=\columnwidth]{figures/H0-P2.pdf}
+    \caption{Posterior on $H_0$ for the mock catalog discussed in \S\,
+    \ref{sec:weakly-variable}.  The blue curve is the posterior when fitting
+    $P_{\detected,\EM}$ with $N_l = 2$.  The fitted value is $H_0 =
+    \variable{output/H0-P2.txt}$ (median and 68\% credible interval).  The
+    orange curve is the posterior ignoring inclination-dependent EM selection
+    effects, which results in $H_0 = \variable{output/H0-P2-nofit.txt}$, a $\sim
+    2\sigma$ bias.  The black vertical line is the true $H_0$ used to generate
+    the catalog.}
+    \label{fig:H0-P2}
+\end{figure}
+
+\subsection{Fitting a Strongly Variable EM Selection}
+\label{sec:strongly-variable}
+
+Different EM counterparts will exhibit different dependence on inclination
+angle.  In this subsection, we model an GW+EM catalog composed of two
+components.  One component has EM detectable emission only very close to on/off
+axis, with $P_{\detected,\EM} = 0$ unless $|x| \geq x_\mathrm{min}$ with
+$x_\mathrm{min} = \cos 15^\circ$.  The other component has a (weak) preference
+for on/off axis emission, with detectability varying quadratically from a
+maximum at $|x| = 1$ to and 50\% of its maximum value at $x = 0$.  The two
+components are mixed equally in the data set, so that the overall EM detection
+efficiency is 
+\begin{equation}
+    P_{\detected,\EM}(x) \propto \frac{1}{2} \left( \frac{H\left( |x| - x_\mathrm{min} \right)}{x_\mathrm{min}} + \frac{3 \left( x^2 + 1/2 \right)}{5} \right),
+\end{equation}
+with $H$ the Heaviside step function.  We choose $N_l = 4$.  The logarithm of
+this function is not expressible as any finite sum of Legendre polynomials, so
+our model is incapable of exactly reproducing the shape of this detection
+efficiency.  Nonetheless, we will see that we can still obtain sub-percent
+accurate, un-biased estimates of $H_0$.
+
+Once again, we generate a catalog of $N = 128$ GW+EM mock detections with S/N
+$\rho > 10$ using $H_0 = 0.7$.  We recover a $P_{\detected,\EM}$ shape that is
+indicative of the EM detection efficiency, but the true efficiency lies well
+outside the posterior 95\% credible interval over much of the range $-1 < x <
+1$.  Figure \ref{fig:top-hat-Pdet} shows the posterior on $P_{\detected,\EM}$.
+
+\begin{figure}
+    \script{top-hat-Pdet.py}
+    \includegraphics[width=\columnwidth]{figures/top-hat-Pdet.pdf}
+    \caption{Posterior on $P_{\detected,\EM}$ from fitting the mock catalog
+    described in \S\, \ref{sec:strongly-variable}.  The catalog is an equal
+    mixture of an on/off axis component ($|x| > \cos 15^\circ$) and a
+    quadratically-varying component; the true $P_{\detected,\EM}$ is shown by
+    the black curve.  The blue line shows the posterior median, and the dark and
+    light bands the posterior 68\% and 95\% credible intervals.  For most of the
+    interval $-1 < x < 1$ the true detection efficiency is well outside the
+    posterior 95\% credible interval due to the mismatch between the Legendre
+    polynomial fitting function, Eq.\ \eqref{eq:Pdet-model} with $N_l = 4$, and
+    the sharply-varying detection efficiency.}
+    \label{fig:top-hat-Pdet}
+\end{figure}
+
+Even though the posterior on the detection efficiency does favors curves very
+different to the true detection efficiency, the posterior on $H_0$ is not
+biased.  The imperfect measurements of inclination from the GW data ``soften''
+the bias in the recovery of the detection efficiency enough to recover the a
+reasonable posterior on $H_0$.  This is shown in Figure \ref{fig:H0-top-hat},
+which also shows that failing to account for the inclination-dependent selection
+effects here leads to very significant bias in the recovered $H_0$ value.
+
+\begin{figure}
+    \script{H0-top-hat.py}
+    \includegraphics[width=\columnwidth]{figures/H0-top-hat.pdf}
+    \caption{Posterior on $H_0$ from fitting the mock catalog described in \S\,
+    \ref{sec:strongly-variable}.  The vertical black line shows $H_0 = 0.7$, the
+    value used to prepare the catalog.  The blue curve shows the posterior when
+    fitting $P_{\detected,\EM}$ using Eq.\ \eqref{eq:Pdet-model} with $N_l=4$.
+    In this case we find $H_0 = \variable{output/H0-top-hat.txt}$ (median and
+    68\% credible interval).  The orange curve shows the posterior when fitting
+    without accounting for inclination-dependent EM selection.  In this case we
+    find $H_0 = \variable{output/H0-top-hat-flat.txt}$, with $\sim 8 \sigma$
+    bias.}
+    \label{fig:H0-top-hat}
 \end{figure}
@@
 \begin{acknowledgments}
-    We thank [people] for conversations.
+    \todo{We thank people for stuff.}
 \end{acknowledgments}
@@
 
+\clearpage
+
 \bibliography{bib}

### 2022-10-01 src/scripts/H0-P2.py

@@ -1,0 +1,26 @@
+import arviz as az
+import matplotlib.pyplot as plt
+import numpy as np
+import os.path as op
+import paths
+import seaborn as sns
+
+sns.set_theme('paper', 'ticks')
+
+trace = az.from_netcdf(op.join(paths.data, "Pl_2_chains.nc"))
+trace_fixed = az.from_netcdf(op.join(paths.data, "Pl_2_fixed_chains.nc"))
+trace_nofit = az.from_netcdf(op.join(paths.data, 'Pl_2_no_fit_chains.nc'))
+
+dim = ['chain', 'draw']
+
+sns.kdeplot(trace.posterior.h.stack(dim=dim), label=r'Fit $P_{\mathrm{det},\mathrm{EM}}$')
+sns.kdeplot(trace_nofit.posterior.h.stack(dim=dim), label=r'Ignore $P_{\mathrm{det},\mathrm{EM}}$')
+
+plt.axvline(0.7, color='k')
+
+plt.xlabel(r'$H_0$')
+plt.ylabel(r'$p\left( H_0 \right)$')
+plt.legend()
+plt.tight_layout()
+
+plt.savefig(op.join(paths.figures, 'H0-P2.pdf'))

### 2022-10-01 src/scripts/H0-top-hat.py

@@ -1,0 +1,26 @@
+import arviz as az
+import matplotlib.pyplot as plt
+import numpy as np
+import os.path as op
+import paths
+import seaborn as sns
+
+sns.set_theme('paper', 'ticks')
+
+trace_hat = az.from_netcdf(op.join(paths.data, 'top_hat_chains.nc'))
+trace_nofit = az.from_netcdf(op.join(paths.data, 'top_hat_flat_inclination_chains.nc'))
+
+dim = ['chain', 'draw']
+
+sns.kdeplot(trace_hat.posterior.h.stack(dim=dim), label=r'Fit $P_{\mathrm{det},\mathrm{EM}}$')
+sns.kdeplot(trace_nofit.posterior.h.stack(dim=dim), label='Ignore $P_{\mathrm{det},\mathrm{EM}}$')
+
+plt.axvline(0.7, color='k')
+
+plt.legend()
+
+plt.xlabel(r'$H_0$')
+plt.ylabel(r'$p\left( H_0 \right)$')
+
+plt.tight_layout()
+plt.savefig(op.join(paths.figures, 'H0-top-hat.pdf'))

### 2022-10-01 src/scripts/inclination-P2.py

@@ -1,0 +1,30 @@
+import arviz as az
+import matplotlib.pyplot as plt
+import numpy as np
+import os.path as op
+import seaborn as sns
+import sys
+import paths
+
+sns.set_theme(context='paper', style='ticks')
+
+trace = az.from_netcdf(op.join(paths.data, 'Pl_2_chains.nc'))
+
+d = ['chain', 'draw']
+
+pd = trace.posterior.P_det
+x = pd.coords['x']
+l, = plt.plot(x, pd.median(dim=d), label='Fit')
+plt.fill_between(x, pd.quantile(0.84, dim=d), pd.quantile(0.16, dim=d), color=l.get_color(), alpha=0.25)
+plt.fill_between(x, pd.quantile(0.975, dim=d), pd.quantile(0.025, dim=d), color=l.get_color(), alpha=0.25)
+
+pd_true = np.exp(0.5*(3*x*x-1)) # This is a hack---should really include models.py
+pd_true = pd_true / np.trapz(pd_true, x)
+plt.plot(x, pd_true, color='k', label='Truth')
+
+plt.xlabel(r'$x$')
+plt.ylabel(r'$P_{\mathrm{det},\mathrm{EM}}\left( x \right)$')
+plt.legend()
+plt.tight_layout()
+
+plt.savefig(op.join(paths.figures, 'inclination-P2.pdf'))

### 2022-10-01 src/scripts/top-hat-Pdet.py

@@ -1,0 +1,38 @@
+import arviz as az
+import matplotlib.pyplot as plt
+import numpy as np
+import os.path as op
+import paths
+import seaborn as sns
+
+sns.set_theme('paper', 'ticks')
+
+# Ugh---pasted from notebook.
+opening_angle = 15
+xmin = np.cos(opening_angle*np.pi/180)
+top_hat_fraction = 0.9
+def Pdet_top_hat(x):
+    x = np.atleast_1d(x)
+    p_broad = 3*(x*x + 0.5)/5
+    p_narrow = np.where(np.abs(x)>xmin, 0.5/(1-xmin), 0)
+
+    return (1-top_hat_fraction)*p_broad+top_hat_fraction*p_narrow
+
+dim = ['chain', 'draw']
+
+trace = az.from_netcdf(op.join(paths.data, 'top_hat_chains.nc'))
+pd = trace.posterior.P_det
+x = pd.coords['x']
+
+l, = plt.plot(x, pd.median(dim=dim), label='Fit')
+plt.fill_between(x, pd.quantile(0.84, dim=dim), pd.quantile(0.16, dim=dim), alpha=0.25, color=l.get_color())
+plt.fill_between(x, pd.quantile(0.975, dim=dim), pd.quantile(0.025, dim=dim), alpha=0.25, color=l.get_color())
+
+plt.plot(x, Pdet_top_hat(x), color='k', label='Truth')
+
+plt.xlabel(r'$x$')
+plt.ylabel(r'$P_{\mathrm{det},\mathrm{EM}}(x)$')
+plt.legend()
+
+plt.tight_layout()
+plt.savefig(op.join(paths.figures, 'top-hat-Pdet.pdf'))

### 2022-10-01 src/scripts/variables.py

@@ -1,0 +1,26 @@
+import arviz as az
+import os.path as op
+import paths
+
+dim = ['chain', 'draw']
+
+def H0_range(file):
+    trace = az.from_netcdf(op.join(paths.data, file))
+    h = trace.posterior.h.stack(dim=dim)
+    return h.median(), h.quantile(0.84), h.quantile(0.16)
+def H0_format(m, h, l):
+    return '{:.4f}^{{+{:.4f}}}_{{-{:.4f}}}'.format(m, h-m, m-l)
+    
+
+# H0 from P2
+with open(op.join(paths.output, 'H0-P2.txt'), 'w') as out:
+    out.write(H0_format(*H0_range('Pl_2_chains.nc')))
+
+with open(op.join(paths.output, 'H0-P2-nofit.txt'), 'w') as out:
+    out.write(H0_format(*H0_range('Pl_2_no_fit_chains.nc')))
+
+# H0 from top hat
+with open(op.join(paths.output, 'H0-top-hat.txt'), 'w') as out:
+    out.write(H0_format(*H0_range('top_hat_chains.nc')))
+with open(op.join(paths.output, 'H0-top-hat-flat.txt'), 'w') as out:
+    out.write(H0_format(*H0_range('top_hat_flat_inclination_chains.nc')))

### 2022-10-01 src/tex/ms.tex
Adding ack'n for KICP workshop.
Fix some typos, lower y-axis limits.

@@
 \affiliation{\cca}
+
+\author{Friends}
 
@@
 for on/off axis emission, with detectability varying quadratically from a
-maximum at $|x| = 1$ to and 50\% of its maximum value at $x = 0$.  The two
+maximum at $|x| = 1$ to and 1/3 of its maximum value at $x = 0$.  The two
 components are mixed equally in the data set, so that the overall EM detection
@@
 our model is incapable of exactly reproducing the shape of this detection
-efficiency.  Nonetheless, we will see that we can still obtain sub-percent
-accurate, un-biased estimates of $H_0$.
+efficiency for any choice of $N_l$.  Nonetheless, we will see that we can still
+obtain sub-percent accurate, un-biased estimates of $H_0$, even with only four
+additional parameters to describe the EM selection function.
 
@@
 \begin{acknowledgments}
-    \todo{We thank people for stuff.}
+    \todo{We thank people for stuff.}  This work was begun at the KICP workshop
+    ``The Quest for Precision Gravitational Wave Cosmology'' in September 2022;
+    the authors thank the KICP and workshop organizers for an exceptionally
+    stimulating environment.
 \end{acknowledgments}

### 2022-10-01 src/scripts/top-hat-Pdet.py

@@ -34,5 +34,7 @@ plt.xlabel(r'$x$')
 plt.ylabel(r'$P_{\mathrm{det},\mathrm{EM}}(x)$')
 plt.legend()
 
+plt.ylim(bottom=0)
+
 plt.tight_layout()
 plt.savefig(op.join(paths.figures, 'top-hat-Pdet.pdf'))

### 2022-10-03 src/tex/ms.tex
Compare structured emission to isotropic (with more sources).

@@
 maximum at $|x| = 1$ to and 1/3 of its maximum value at $x = 0$.  The two
-components are mixed equally in the data set, so that the overall EM detection
-efficiency is 
+components are mixed equally in the population of binary mergers, so that the
+overall EM detection efficiency is 
 \begin{equation}
@@
     without accounting for inclination-dependent EM selection.  In this case we
-    find $H_0 = \variable{output/H0-top-hat-flat.txt}$, with $\sim 8 \sigma$
+    find $H_0 = \variable{output/H0-top-hat-flat.txt}$, with $\sim 6 \sigma$
     bias.}
     \label{fig:H0-top-hat}
+\end{figure}
+
+With this choice of EM detection efficiency, we observe a counterpart in about
+15\% of the GW-detectable systems.  It is interesting to ask whether the
+improvement in inclination estimation from the structured EM emission (and
+therefore improved distance estimation from the GW observations) outweighs the
+loss of sources compared to a scenario where the EM emission is uniform.  (In
+other words, would it be better to employ a search strategy that selects sources
+that are likely to have anisotropic EM emission in order to improve distance
+estimates?)  For this distribution, a catalog with $1/0.15 \sim 7$ times as many
+sources with isotropic EM emission provides overall a better Hubble parameter
+measurement, as shown in Figure \ref{fig:top-hat-flat-compare}.  (For both mock
+catalogs, we employ $N_l = 4$, allowing for the same possibility of structure in
+the EM emission; the inferred $P_{\detected,\EM}$ for the catalog with isotropic
+emission is shown in Figure \ref{fig:flat-Pdet}.)
+
+\begin{figure}
+    \script{top-hat-flat-compare.py}
+    \includegraphics[width=\columnwidth]{figures/top-hat-flat-compare.pdf}
+    \caption{Comparison between $H_0$ inferred from a mock catalog with the
+    anisotropic EM emission discussed in \S\ \ref{sec:strongly-variable} (blue
+    curve) with a catalog with an equivalent GW merger rate but isotropic
+    emission (and therefore $\sim 7$ times as many detections; orange curve).
+    The improved catalog size from the isotropic emission overcomes the improved
+    distance estimates from the knowledge of inclination gained from the
+    anisotropic emission, and the overall measurement is tighter.  We find the
+    posterior median and 68\% credible interval $H_0 =
+    \variable{output/H0-top-hat.txt}$ from the catalog with anisotropic EM
+    emission, and $H_0 = \variable{output/H0-flat-comparison.txt}$ from the
+    larger catalog with isotropic emission.}
+    \label{fig:top-hat-flat-compare}
+\end{figure}
+
+\begin{figure}
+    \script{flat-Pdet.py}
+    \includegraphics[width=\columnwidth]{figures/flat-Pdet.pdf}
+    \caption{The inferred inclination distribution for the catalog with
+    isotropic EM emission and equivalent GW merger to the anisotropic emission
+    described in \S\ \ref{sec:strongly-variable}.  The blue line gives the
+    posterior median and the dark and light bands the posterior 68\% and 95\%
+    credible interval.  Even though the EM emission is isotropic, we fit the
+    detection efficiency with $N_l = 4$ for a fair comparison with the (smaller)
+    catalog of anisotropic sources whose inferred inclination distribution is
+    shown in Figure \ref{fig:top-hat-Pdet}.}
+    \label{fig:flat-Pdet}
 \end{figure}
@@
 
-\software{\texttt{arviz} \citep{arviz_2019}, \texttt{matplotlib} \citep{Hunter:2007}, \texttt{pymc} \citep{salvatier2016probabilistic}, \texttt{seaborn} \citep{Waskom2021}, \texttt{showyourwork} \citep{Luger2021}}
+\software{\texttt{arviz} \citep{arviz_2019}, \texttt{matplotlib}
+\citep{Hunter:2007}, \texttt{pymc} \citep{salvatier2016probabilistic},
+\texttt{seaborn} \citep{Waskom2021}, \texttt{showyourwork} \citep{Luger2021}}
 

### 2022-10-03 src/notebooks/models.py

@@ -50,7 +50,6 @@ def draw_gws(P_det, N, fr=1.0, fl=1.0, h=0.7, dmax=0.5, rho_thresh=10):
     Al = []
     Ar = []
 
-    N = 2048
     Ndraw = 0
     with tqdm(total=N) as bar:
         while len(x_true) < N:

### 2022-10-03 src/scripts/flat-Pdet.py

@@ -1,0 +1,29 @@
+import arviz as az
+import matplotlib.pyplot as plt
+import numpy as np
+import os.path as op
+import paths
+import seaborn as sns
+
+sns.set_theme('paper', 'ticks')
+
+trace = az.from_netcdf(op.join(paths.data, 'top_hat_flat_comparison.nc'))
+
+dim = ['chain', 'draw']
+
+pd = trace.posterior.P_det
+x = pd.coords['x']
+
+l, = plt.plot(x, pd.median(dim=dim), label='Fit')
+plt.fill_between(x, pd.quantile(0.84, dim=dim), pd.quantile(0.16, dim=dim), color=l.get_color(), alpha=0.25)
+plt.fill_between(x, pd.quantile(0.975, dim=dim), pd.quantile(0.025, dim=dim), color=l.get_color(), alpha=0.25)
+
+plt.plot(x, 0.5*np.ones_like(x), label='Truth', color='k')
+
+plt.xlabel(r'$x$')
+plt.ylabel(r'$P_{\mathrm{det},\mathrm{EM}}(x)$')
+
+plt.legend()
+
+plt.tight_layout()
+plt.savefig(op.join(paths.figures, 'flat-Pdet.pdf'))

### 2022-10-03 src/scripts/top-hat-flat-compare.py

@@ -1,0 +1,26 @@
+import arviz as az
+import matplotlib.pyplot as plt
+import numpy as np
+import os.path as op
+import paths
+import seaborn as sns
+
+sns.set_theme('paper', 'ticks')
+
+trace_hat = az.from_netcdf(op.join(paths.data, 'top_hat_chains.nc'))
+trace_flat = az.from_netcdf(op.join(paths.data, 'top_hat_flat_comparison.nc'))
+
+dim = ['chain', 'draw']
+
+sns.kdeplot(trace_hat.posterior.h.stack(dim=dim), label=r'Anisotropic')
+sns.kdeplot(trace_flat.posterior.h.stack(dim=dim), label=r'Isotropic')
+
+plt.axvline(0.7, color='k')
+
+plt.legend()
+
+plt.xlabel(r'$H_0$')
+plt.ylabel(r'$p\left( H_0 \right)$')
+
+plt.tight_layout()
+plt.savefig(op.join(paths.figures, 'top-hat-flat-compare.pdf'))

### 2022-10-03 src/scripts/variables.py

@@ -24,3 +24,7 @@ with open(op.join(paths.output, 'H0-top-hat.txt'), 'w') as out:
     out.write(H0_format(*H0_range('top_hat_chains.nc')))
 with open(op.join(paths.output, 'H0-top-hat-flat.txt'), 'w') as out:
     out.write(H0_format(*H0_range('top_hat_flat_inclination_chains.nc')))
+
+# Comparison from top hat versus flat catalog
+with open(op.join(paths.output, 'H0-flat-comparison.txt'), 'w') as out:
+    out.write(H0_format(*H0_range('top_hat_flat_comparison.nc')))

### 2022-10-06 src/tex/ms.tex
Calculate how much posterior improves when we *know* P_det.

@@
     \ref{sec:weakly-variable}.  The blue curve is the posterior when fitting
-    $P_{\detected,\EM}$ with $N_l = 2$.  The fitted value is $H_0 =
+    $P_{\detected,\EM}$ with $N_l = 2$.  We find $H_0 =
     \variable{output/H0-P2.txt}$ (median and 68\% credible interval).  The
-    orange curve is the posterior ignoring inclination-dependent EM selection
-    effects, which results in $H_0 = \variable{output/H0-P2-nofit.txt}$, a $\sim
-    2\sigma$ bias.  The black vertical line is the true $H_0$ used to generate
-    the catalog.}
+    orange curve is the posterior when $P_{\detected,\EM}$ is fixed to its true
+    value (i.e.\ we assume we know the EM selection effect perfectly).  We find
+    $H_0 = \variable{output/H0-P2-fixed.txt}$, a 5 percent reduction in
+    uncertainty.  The green curve is the posterior ignoring
+    inclination-dependent EM selection effects, which results in $H_0 =
+    \variable{output/H0-P2-nofit.txt}$, a $\sim 2\sigma$ bias.  The black
+    vertical line is the true $H_0$ used to generate the catalog.}
     \label{fig:H0-P2}

### 2022-10-06 src/scripts/H0-P2.py

@@ -14,6 +14,7 @@ trace_nofit = az.from_netcdf(op.join(paths.data, 'Pl_2_no_fit_chains.nc'))
 dim = ['chain', 'draw']
 
 sns.kdeplot(trace.posterior.h.stack(dim=dim), label=r'Fit $P_{\mathrm{det},\mathrm{EM}}$')
+sns.kdeplot(trace_fixed.posterior.h.stack(dim=dim), label=r'Fix $P_{\mathrm{det},\mathrm{EM}}$')
 sns.kdeplot(trace_nofit.posterior.h.stack(dim=dim), label=r'Ignore $P_{\mathrm{det},\mathrm{EM}}$')
 
 plt.axvline(0.7, color='k')

### 2022-10-06 src/scripts/variables.py

@@ -16,6 +16,9 @@ def H0_format(m, h, l):
 with open(op.join(paths.output, 'H0-P2.txt'), 'w') as out:
     out.write(H0_format(*H0_range('Pl_2_chains.nc')))
 
+with open(op.join(paths.output, 'H0-P2-fixed.txt'), 'w') as out:
+    out.write(H0_format(*H0_range('Pl_2_fixed_chains.nc')))
+
 with open(op.join(paths.output, 'H0-P2-nofit.txt'), 'w') as out:
     out.write(H0_format(*H0_range('Pl_2_no_fit_chains.nc')))
 
